---
title: "Intermediate elective"
author: "Audrey Juneman"
format: pdf
date: "2026-05-15"
---

# Set up

## Packages

```{r}
#| label: packages

# general use and cleaning
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrapping axis labels
library(scales)

# reading in .xlsx files
library(readxl)

# visualizing missing data
library(naniar)

# arranging plots
library(patchwork)

#colors 
library(NatParksPalettes)

#copied from online code 

suppressPackageStartupMessages({
pacman::p_load(
  tidyverse, ggtext, showtext, 
  janitor, scales, glue
  )
})

### |- figure size ----
camcorder::gg_record(
  dir    = here::here("temp_plots"),
  device = "png",
  width  = 6,
  height = 10,
  units  = "in",
  dpi    = 320
)

install.packages("rstudioapi")
library(rstudioapi)
current_theme <- getThemeInfo()
print(current_theme)

install.packages("devtools")
```

## Data 

```{r}
#| label: data 

veg <- read_csv(here("data", "veg.csv"))

metadata <- read_csv(here("data", "vp_veg_metadata.csv"))

# create a new object from veg
pools_implicitNA <- veg |> 
  # filter all occurrences where transect_name contains VP
  filter(str_detect(transect_name, "VP")) |> 
  # group by year, pool, and species
  group_by(year, transect_name, psoc) |>
  # calculate mean percent cover
  summarize(mean_pc = mean(percent_cover, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |>
  # join with metadata by year_pool
  left_join(metadata, by = "year_pool")

# create a new object from veg
pools_explicit0 <- veg |> 
  # filter all occurrences where transect_name contains VP
  filter(str_detect(transect_name, "VP")) |> 
  # complete all possible combinations of year, pool, psoc and distance
  complete(year, transect_name, psoc, transect_distance,
           # fill values of percent cover with 0
           fill = list(percent_cover = 0)
           ) |>
  # group by year, pool, and psoc
  group_by(year, transect_name, psoc) |>
  # calculate total percent cover across all quads
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
  # ungroup the data frame
  ungroup() |>
  # create a new column called year_pool from year and transect name
  unite("year_pool", year, transect_name, remove = TRUE) |> 
  # join with metadata data frame
  left_join(metadata, by = "year_pool") |> 
  # calculate mean percent cover from total percent cover/number of quadrats
  mutate(mean_pc = sum_pc/num_quad)

# creating new ojbect from implicit NA data frame
filtered_implicitNA <- pools_implicitNA |> 
  # filter to only include 
  filter(psoc == "Eryngium_vaseyi")

filtered_explicit0 <- pools_explicit0 |> 
  filter(psoc == "Eryngium_vaseyi")

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# aquatic invertebrates
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality information
water_quality <- read_xlsx(here("data",
                                "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 

# creating new object from sites
ncos_sites <- sites |> 
  # filter to only include NCOS sites sampled four times a year
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")

# creating new clean object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converting taxon name to snake case (no spaces or capital letters,
  # only underscores)
  mutate(verbatim_name = to_any_case(verbatim_name, case = "snake"))

# creating new clean object from water quality
water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |> 
  # calculate median pH, salinity, DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers
  filter(date_site != "2022-08-12_NVBR") 

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance") |> 
  # group by date, site, taxon
  group_by(date, site, taxon) |> 
  # calculate average abundance per liter 
  # (sum abundance divided by 7.5 liters)
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  # ungroup data frame
  ungroup() |> 
  # create a new column called `date_site` to join with water quality
  unite("date_site", date, site, remove = FALSE) |> 
  # join with water quality
  left_join(water_quality_clean, by = c("date_site")) |> 
  # join with taxon list
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  # add full names of sites
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  # setting factor levels for site
  mutate(site_full = fct_reorder(site_full, med_sal,
                                 .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))

```

#1. Explain your inspiration

- For my Intermediate elective, I chose the coal falls and renewables rise visualization, due to my interest in energy usage in the United States. I enjoy the color scheme and how easy to interpret the graph is due to the nice colors and organization. 
- I think this makes sense for comparing ostracods and copepod abundance throughout time, because it compares two specific variables and shows their trends throughout time. I will specificlaly be looking at Devereaux creek. 
- My x axis will be `date_site` and my y-axis will be `taxon`. It will be colored by `taxon`. 

#2. Plan your figure 

#3. Code your figure 

```{r}
#| label: filtering and tidying data

#filtering dataset for copepod line
aq_ins_dev <- aq_ins_clean |>  
  filter(site_full == "Devereux Creek")
```

```{r}
#| label: code-to-create-color-function

#' Get default theme colors
#' @param palette Optional custom color palette
#' @return List of theme colors
get_theme_colors <- function(palette = NULL) {
    list(
        background = "#FDFDFD",      #"#f5f5f2"
        title      = "gray20",
        subtitle   = "gray30",
        text       = "gray30",
        caption    = "gray40",
        palette    = palette  # This can be overridden weekly
    )
}

#' Create base theme with consistent elements
#' @param colors List of colors from get_theme_colors()
#' @return ggplot theme object
create_base_theme <- function(colors = get_theme_colors()) {
    # Only include the truly fixed elements that don't change weekly
    theme_minimal(base_size = 14, base_family = fonts$text) +
        theme(
            plot.title.position   = "plot",
            plot.caption.position = "plot",
            legend.position       = "plot",
            
            # Background elements
            plot.background  = element_rect(fill = colors$background, color = colors$background),
            panel.background = element_rect(fill = colors$background, color = colors$background),
            
            # Common margins
            plot.margin = margin(t = 10, r = 20, b = 10, l = 20),
            
            # Text elements that stay consistent
            axis.text  = element_text(
                family = fonts$text,
                size   = rel(0.79),
                color  = colors$text
            ),
            axis.title = element_text(
                family = fonts$text,
                size   = rel(0.93),
                face   = "bold",
                color  = colors$text
            )
            
            # Note: Weekly-specific theme elements should be added in the weekly scripts
        )
}

#' Extend base theme with weekly customizations
#' @param base_theme Existing theme to extend
#' @param weekly_theme_elements List of theme elements specific to this week
#' @return Modified ggplot theme
extend_weekly_theme <- function(base_theme, weekly_theme_elements) {
    base_theme + weekly_theme_elements
}
```



```{r}
#| label: visualization-params

### |- plot aesthetics ----
colors <- get_theme_colors(palette = list(
  bg_color     = "#0D1B2A",
  solar_gold   = "#F4A932", 
  coal_blue    = "#6A80A0",   
  text_white   = "#FFFFFF",
  text_muted   = "#8A98A8",
  text_sub     = "#B0B8C4",
  grid_color    = "#1A2E42",
  accent_coral = "#E8503A"   
))

bg_color     <- colors$palette$bg_color
solar_gold   <- colors$palette$solar_gold
coal_blue    <- colors$palette$coal_blue
text_white   <- colors$palette$text_white
text_muted   <- colors$palette$text_muted
text_sub     <- colors$palette$text_sub
grid_color   <- colors$palette$grid_color
accent_coral <- colors$palette$accent_coral
col_renew <- solar_gold
col_coal  <- coal_blue


### |- titles and caption ----
title_text <- glue(
  "<span style='font-weight:900; color:{coal_blue};'>Copepods rise???.&nbsp;</span>",
  "<span style='font-weight:900; color:{solar_gold};'> Ostracods rise??.</span>"
)

subtitle_text <- glue(
  "<span style='font-size:12pt; font-weight:400; color:{text_sub};'>",
  "The relationship between the two species<br>",
  "Over three years · 2020–2023",
  "</span>"
)
```

```{r}
#| label: plotting-graphs 

#| label: plot
#| warning: false

### |- main plot ----
graph <- aq_ins_dev |>
  ggplot(aes(x = date, y = ave_lit, color = taxon)) +

  # Geom
  geom_hline(
    yintercept = seq(0, 80, by = 20),
    color = grid_color,
    linewidth = 0.25
  ) +
  geom_line(
    data = aq_ins_dev |> filter(taxon == "copepod"),
    linewidth = 1.9,
    alpha = 1.0
  ) +
  geom_line(
    data = aq_ins_dev |> filter(taxon == "ostracod"),
    linewidth = 1.3,
    alpha = 0.60
  ) +
  geom_point(
    data = aq_ins_dev |> filter(date == 2020-01-21),
    size = 3.5,
    shape = 21,
    fill = bg_color,
    stroke = 1.8
  ) +
  geom_point(
    data = aq_ins_dev |> filter(date == 2020-01-21),
    size = 3.5,
    shape = 21,
    fill = bg_color,
    stroke = 1.8
  ) +

  # Scales
  scale_color_manual(
    values = c("ostracod" = aq_ins_dev, "copepod" = aq_ins_dev)
  ) +
  scale_x_continuous(
    breaks = c(2020, 2021, 2022, 2023),
    expand = expansion(mult = c(0.03, 0.24))
  ) +
  scale_y_continuous(
    limits = c(0, 90),
    breaks = seq(0, 80, by = 20),
    labels = \(x) glue("{x}%"),
    expand = expansion(mult = c(0.01, 0.04))
  ) +

  # Labs
  labs(
    title = "Copepod and Ostracod abundance in Devereux Creek from 2020-2023",
    x = NULL,
    y = NULL
  ) +

  # Theme
  theme_void() +
  theme(
    # canvas
    plot.background = element_rect(fill = bg_color, color = NA),
    panel.background = element_rect(fill = bg_color, color = NA),
    plot.title = element_markdown(
      family = fonts$title,
      size = 28,
      color = text_white,
      hjust = 0,
      lineheight = 1.1,
      margin = margin(t = 10, b = 6)
    ),
    plot.subtitle = element_markdown(
      family = fonts$text,
      size = 10,
      color = text_sub,
      hjust = 0,
      lineheight = 1.4,
      margin = margin(b = 22)
    ),
    plot.caption = element_markdown(
      family = fonts$text,
      size = 6.5,
      color = "#5A6878",
      hjust = 1,
      margin = margin(t = 14)
    ),
    axis.text.x = element_text(
      family = fonts$text,
      size   = 9,
      color  = text_muted,
      margin = margin(t = 6)
    ),
    axis.text.y = element_text(
      family = fonts$text,
      size = 8,
      color = text_muted,
      hjust = 1
    ),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
    plot.margin = margin(t = 20, r = 16, b = 16, l = 20)
  )




```

